Box

### 2021-07-14 MS/box.Rmd
Fixed equations so they are uniform with MS

---
bibliography: ../bib/refs.bib
csl: ../bib/the-journal-of-experimental-biology.csl
output: html_document
---

@@

$$
\text{ARR} = \frac{X_{1} - X_{2}}{T_2 - T_1},
$$

where *T* stands for temperature in degrees Celsius and $T_{2} > T_{1}$, and $X_{1}$ and $X_{2}$ are the average physiological responses (estimated means) at two temperature points $T_{1}$ and $T_{2}$.

To obtain the sampling variance for this equation (slope), we need to know several basic properties of variance. Let's assume $X_{1}$ is a random variable, which means that this variable has a mean and standard deviation (variance). Multiplying a constant ($a$) will change the variance by the square of that constant ($a^2$) while adding or subtracting the constant ($b$) does not change the variance of $X_{1}$. This can be summarized as:

@@

Also, when adding two random variables ($X_{1}$ and $X_{2}$), the combined variance is the sum of the variance of $X_{1}$  and the variance of $X_{2}$ plus 2 times the covariance between $X_{1}$ and $X_{2}$. This relationship can be written as:

$$
\sigma^2_{X_1 \pm X_2} = \sigma^2_{X_1} + \sigma^2_{X_2} \pm 2\text{Cov}(X_1,X_2) \\
= \sigma^2_{X_1} + \sigma^2_{X_2} \pm 2\text{Cor}(X_1,X_2)\sqrt{ \sigma^2_{X_1} \sigma^2_{X_2} }
$$ 
where the co-variance $\text{Cov}(X_1,X_2)$ equals to the correlation multiplied by the square-root of the two variances $2\text{Cor}(X_1,X_2)\sqrt{ \sigma^2_{X_1} \sigma^2_{X_2} }$.

Importantly, when $X_{1}$ and $X_{2}$ are independent of each other the covariance is 0. More concretely, if measurements are done on two different groups of animals at two different temperature points ($T_1$ and $T_2$), then the co-variances between these two sets of measurements are 0.

Therefore, when $X_{1}$ and $X_{2}$ are independent, we can obtain the sampling variance for ARR as:

@@

Now, it is not uncommon that we use the same group of animals (or plants) at two temperature points. If we do, however, then we need to add the co-variance between $X_{1}$ and $X_{2}$ ($\text{Cov}(X_{1},X_{2})$); note that as above, the co-variance equals the corresponding correlation multiplied by the square-root of two (sampling) variances. Therefore, the sampling variance can be now written as:

$$
\sigma^2_{AAR} = \left( \frac{1} {T_{2} - T_{1}} \right)^2 
\left( 
\frac{SD^2_{1}} {N_1} + \frac{SD^2_{2}} {N_2} - 2r \sqrt{\frac{SD^2_{1}} {N_1}  \frac{SD^2_{2}} {N_2}} 
\right)
$$ 
By assuming the number of animals (*N*) are the same at the two temperature points, we can slightly simplify this formula:

@@

where *r* is the correlation between a set of measurement $X_1$ and $X_2$ form the same individuals at two point $T_1$ and $T_2$. As you may notice, we need raw data to calculate *r*. So, in reality, we often need to assume a certain value of *r*. @Noble2017 discuss why we can assume $r = 0.5$ and this may be the most reasonable thing to do when we do not have any estimate of *r*.

@@

$$
\text{Var}(\text{ARRD}) = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( \frac{SD^2_{f1}} {N_{f1}} + \frac{SD^2_{f2}} {N_{f2}} +
 \frac{SD^2_{m1}} {N_{m1}} + \frac{SD^2_{m2}} {N_{m2}}
\right),
$$ where subscripts *f* and *m* stands for 'females' and 'males'.

@@

$$
\text{Var}(\text{ARRD}) = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( 
\frac{SD^2_{f1} + SD^2_{f2}  - 2rSD_{f1}SD_{f2}} {N_f}   +
\frac{SD^2_{m1} + SD^2_{m2}  - 2rSD_{m1}SD_{m2}} {N_m}
\right).
$$

@@

$$
\text{Var}(\text{ARRD}) = \left( \frac{1} {T_4 - T_3} \right)^2 
\left( 
\frac{SD^2_{f3} + SD^2_{f4}  - 2rSD_{f3}SD_{f4}} {N_f}   +
\right) +
\left( \frac{1} {T_6 - T_5} \right)^2 
\left( 
\frac{SD^2_{m5} + SD^2_{m6}  - 2rSD_{m5}SD_{m6}} {N_m}
\right).
$$

@@

$$
SMD_{T} = \frac{\left(\bar{X}_{1} - \bar{X}_{2}\right)}{SD_{p}\left( T_{2} - T_{1}\right)}J
$$

Here, we can see that the difference between $X_1$ and $X_2$ is standardised by the pooled SD, essentially correcting for unitary differences. Dividing SMD by the temperature difference on top of that results in correcting the effect size further by temperature. To derive the sampling variance for $SMD_T$ we can apply the same principles as we did above to derive an approximate sampling variance as:

@@

$$
\text{Var}(f(X_1)) \approx \text{Var}(X_1)* \left( 
f'(X_1)
\right)^2,
$$ where $f(X_1)$ represents the function of the random variable $X_1$, and importantly, $f'(X_1)$ is the first derivative of $f(X_1)$. Probably it is hard to understand what does this really means without a concrete example. So let's derive the sampling variance for $\ln{RR}$, which is:

$$
\ln{RR} = \ln \left ( \frac{X_{1}}{X_{2}}\right ) = \ln{X_1} -\ln{X_2}.
$$

Here, $f(X_1) = \ln{X_1}$. By applying $f'(X_1) = 1/{X_1}$ (i.e. the first derivative of $lnX1$ is $1/{X_1}$) to the delta method and using the variance's basic properties, we have:

$$
\sigma^2_{ln{RR}} \approx  \left ( \frac{SD_{1}^2}{N_{1}}\right ) * \left(\frac{1}{X_{1}} \right)^2 + \left ( \frac{SD_{2}^2}{N_{2}}\right ) * \left(\frac{1}{X_{2}} \right)^2 =   \left ( \frac{SD_{1}^2}{N_{1}X_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}X_{2}^2}\right )
$$
Note here again that the first term, $\frac{SD_{1}^2}{N_{1}}$ is the sampling variance for group 1, and we simply multiply this term by the square of the first derivative of $X_{1}$ as the Delta method tells us. Of course, readers will see that this is equivalent to the sampling variance for $ln_{RR}$ (Table 1). We can extend this sampling variance to include dependency between the groups as:

$$
\sigma^2_{ln{RR}} \approx  \left ( \frac{SD_{1}^2}{N_{1}X_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}X_{2}^2}\right ) 
- 2r\sqrt{\frac{SD_{1}^2}{N_{1}X_{1}^2}}\sqrt{\frac{SD_{2}^2}{N_{2}X_{2}^2}},
$$
where $r$ is 0 when two sets of measurements are independent (therefore, the term including $r$ disappears), while $r$ is the correlation between $\ln{X_1}$ and $\ln{X_2}$. As discussed, we usually do not have an estimate for this correlation coefficient.

>>>>

---
bibliography: ../bib/refs.bib
csl: ../bib/the-journal-of-experimental-biology.csl
output: 
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    toc: no
    toc_depth: 6
    toc_float: yes
    bibliography: refs.bib
  bookdown::word_document2:
    toc: no
    toc_depth: 6
    number_sections: false
    reference_docx: template.docx
editor_options: 
  chunk_output_type: console
---

@@

$$
\text{ARR} = \frac{M_{1} - M_{2}}{T_2 - T_1},
$$

where *T* stands for temperature in degrees Celsius and $T_{2} > T_{1}$, and $M_{1}$ and $M_{2}$ are the average physiological responses (estimated means) at two temperature points $T_{1}$ and $T_{2}$.

To obtain the sampling variance for this equation (slope), we need to know several basic properties of variance. Let's assume $M_{1}$ is a random variable, which means that this variable has a mean and standard deviation (variance). Multiplying a constant ($a$) will change the variance by the square of that constant ($a^2$) while adding or subtracting the constant ($b$) does not change the variance of $M_{1}$. This can be summarized as:

@@

Also, when adding two random variables ($M_{1}$ and $M_{2}$), the combined variance is the sum of the variance of $X_{1}$  and the variance of $X_{2}$ plus 2 times the covariance between $M_{1}$ and $M_{2}$. This relationship can be written as:

$$
\sigma^2_{M_1 \pm M_2} = \sigma^2_{M_1} + \sigma^2_{M_2} \pm 2\text{Cov}(M_1,M_2) 
= \sigma^2_{M_1} + \sigma^2_{M_2} \pm 2\text{Cor}(M_1,M_2)\sqrt{ \sigma^2_{M_1} \sigma^2_{M_2} }
$$ 
where the co-variance $\text{Cov}(M_1,M_2)$ equals to the correlation multiplied by the square-root of the two variances $2\text{Cor}(M_1,M_2)\sqrt{ \sigma^2_{M_1} \sigma^2_{M_2} }$.

Importantly, when $M_{1}$ and $M_{2}$ are independent of each other the covariance is 0. More concretely, if measurements are done on two different groups of animals at two different temperature points ($T_1$ and $T_2$), then the co-variances between these two sets of measurements are 0.

Therefore, when $M_{1}$ and $M_{2}$ are independent, we can obtain the sampling variance for ARR as:

@@

Now, it is not uncommon that we use the same group of animals (or plants) at two temperature points. If we do, however, then we need to add the co-variance between $M_{1}$ and $M_{2}$ ($\text{Cov}(M_{1},M_{2})$); note that as above, the co-variance equals the corresponding correlation multiplied by the square-root of two (sampling) variances. Therefore, the sampling variance can be now written as:

$$
\sigma^2_{AAR} = \left( \frac{1} {T_{2} - T_{1}} \right)^2 
\left( 
\frac{SD^2_{1}} {N_1} + \frac{SD^2_{2}} {N_2} - 2r \sqrt{\frac{SD^2_{1}} {N_1}  \frac{SD^2_{2}} {N_2}} 
\right)
$$

By assuming the number of animals (*N*) are the same at the two temperature points, we can slightly simplify this formula:

@@

where *r* is the correlation between a set of measurement $M_1$ and $M_2$ form the same individuals at two point $T_1$ and $T_2$. As you may notice, we need raw data to calculate *r*. So, in reality, we often need to assume a certain value of *r*. @Noble2017 discuss why we can assume $r = 0.5$ and this may be the most reasonable thing to do when we do not have any estimate of *r*.

@@

$$
\sigma^2_{ARRD} = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( \frac{SD^2_{f1}} {N_{f1}} + \frac{SD^2_{f2}} {N_{f2}} +
 \frac{SD^2_{m1}} {N_{m1}} + \frac{SD^2_{m2}} {N_{m2}}
\right),
$$

where subscripts *f* and *m* stands for 'females' and 'males'.

@@

$$
\sigma^2_{ARRD} = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( 
\frac{SD^2_{f1} + SD^2_{f2}  - 2rSD_{f1}SD_{f2}} {N_f}   +
\frac{SD^2_{m1} + SD^2_{m2}  - 2rSD_{m1}SD_{m2}} {N_m}
\right).
$$

@@

$$
\sigma^2_{ARRD} = \left( \frac{1} {T_4 - T_3} \right)^2 
\left( 
\frac{SD^2_{f3} + SD^2_{f4}  - 2rSD_{f3}SD_{f4}} {N_f}   +
\right) +
\left( \frac{1} {T_6 - T_5} \right)^2 
\left( 
\frac{SD^2_{m5} + SD^2_{m6}  - 2rSD_{m5}SD_{m6}} {N_m}
\right).
$$

@@

$$
SMD_{T} = \frac{\left(M_{1} - M_{2}\right)}{SD_{p}\left( T_{2} - T_{1}\right)}J
$$

Here, we can see that the difference between $M_1$ and $M_2$ is standardised by the pooled SD, essentially correcting for unitary differences. Dividing SMD by the temperature difference on top of that results in correcting the effect size further by temperature. To derive the sampling variance for $SMD_T$ we can apply the same principles as we did above to derive an approximate sampling variance as:

@@

$$
\text{Var}(f(M_1)) \approx \text{Var}(M_1)* \left( 
f'(M_1)
\right)^2,
$$

where $f(M_1)$ represents the function of the random variable $X_1$, and importantly, $f'(M_1)$ is the first derivative of $f(M_1)$. Probably it is hard to understand what does this really means without a concrete example. So let's derive the sampling variance for $\ln{RR}$, which is:

$$
\ln{RR} = \ln \left ( \frac{M_{1}}{M_{2}}\right ) = \ln{M_1} -\ln{M_2}.
$$

Here, $f(M_1) = \ln{M_1}$. By applying $f'(M_1) = 1/{M_1}$ (i.e. the first derivative of $lnM_1$ is $1/{M_1}$) to the delta method and using the variance's basic properties, we have:

$$
\sigma^2_{ln{RR}} \approx  \left ( \frac{SD_{1}^2}{N_{1}}\right ) * \left(\frac{1}{M_{1}} \right)^2 + \left ( \frac{SD_{2}^2}{N_{2}}\right ) * \left(\frac{1}{M_{2}} \right)^2 =   \left ( \frac{SD_{1}^2}{N_{1}M_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}M_{2}^2}\right )
$$
Note here again that the first term, $\frac{SD_{1}^2}{N_{1}}$ is the sampling variance for group 1, and we simply multiply this term by the square of the first derivative of $X_{1}$ as the Delta method tells us. Of course, readers will see that this is equivalent to the sampling variance for $ln_{RR}$ (Table 1). We can extend this sampling variance to include dependency between the groups as:

$$
\sigma^2_{ln{RR}} \approx  \left ( \frac{SD_{1}^2}{N_{1}M_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}M_{2}^2}\right ) 
- 2r\sqrt{\frac{SD_{1}^2}{N_{1}M_{1}^2}}\sqrt{\frac{SD_{2}^2}{N_{2}M_{2}^2}},
$$
where $r$ is 0 when two sets of measurements are independent (therefore, the term including $r$ disappears), while $r$ is the correlation between $\ln{M_1}$ and $\ln{M_2}$. As discussed, we usually do not have an estimate for this correlation coefficient.
